Minimal trap spaces of Logical models are maximal siphons of their Petri net encoding

### 2022-05-12 cmsb22.tex
Still going…

  A set of places of Petri net \(P\) encoding logical model \(\mathcal{M}\) is \emph{conflict-free} if it does not contain any two places corresponding to the active and inactive states of the same \emph{gene} of \(\mathcal{M}\). Then, a conflict-free siphon \(S\) is said to be \emph{maximal} if and only if there is no other conflict-free siphon \(S'\) such that \(S \subset S'\).

@@

For a few years now the second author has been collaborating with biologists who have been building very large detailed and annotated maps and now wish to analyze the dynamics of the corresponding models.
One of the main maps studied this way is a map representing knowledge about the Rheumatoïd Arthritis~\cite{singh2018computational}, which was the main motivation for the development of a tool to automatically transform it into an executable Boolean model~\cite{aghamiri2020automated}.
In the supplementary material of the paper, an excerpt of the map, focused around the apoptosis (cell death) module is transformed into a model of \emph{reasonable} size, namely 181 Boolean variables (model \verb|F5_RA_apoptosis_executable_module.sbml| of supplementary material S3, and model ``RA-apoptosis'' of Section~\ref{sec:eval}).
The study of such model, though, is a big hurdle.
Indeed, as stated in the article about another model of the same size:
\emph{``The size of the CaSQ-inferred MAPK model (181 nodes) made the calculation of stable states a non-realistic endeavour.''}

@@

However, using a state-of-the-art tool like PyBoolNet~\cite{klarner2015computing} on that model actually fails at the phase of prime-implicant generation.
And hence, it is not possible to extract any (complex) attractor at all.
This is also true for the Alzheimer model also mentioned in that same article and originally from~\cite{ogishima2016alzpathway} (\verb|F4| file in the original supplementary material, and ``Alzheimer'' in Table~\ref{tab:result_real}), but actually not for the MAPK model for which the first trap spaces can be obtained in reasonable time.
The current practice usually revolves then around fixing some inputs to plausible values and reducing the model accordingly.
While this approach makes sense, it relies on potentially arbitrary decisions, and \emph{hides away} critical modelling choices that were actually not part of the original Boolean model.

@@

To assess the efficiency of the proposed method, we compared it with the state-of-the-art method PyBoolNet~\cite{klarner2015computing,klarner2017pyboolnet} on both its own repository of models and large models from the literature.

@@

All the models and a CoLoMoTo notebook realizing the benchmarks can be found at \url{https://github.com/soli/trap-spaces-as-siphons}. These can be run on a Docker image in the cloud by clicking the Binder button.

>>>>

  A set of places of Petri net \(\mathcal{P}\) encoding logical model \(\mathcal{M}\) is \emph{conflict-free} if it does not contain any two places corresponding to the active and inactive states of the same \emph{gene} of \(\mathcal{M}\). Then, a conflict-free siphon \(S\) is said to be \emph{maximal} if and only if there is no other conflict-free siphon \(S'\) such that \(S \subset S'\).

@@

For a few years now we have been collaborating with biologists who build very large detailed and annotated maps and now wish to analyze the dynamics of the corresponding models.
One of the main maps studied this way represents knowledge about the Rheumatoïd Arthritis~\cite{singh2018computational}, and was the main motivation for the development of a tool to automatically transform it into an executable Boolean model~\cite{aghamiri2020automated}.
In the supplementary material of the paper, an excerpt of the map, focused around the apoptosis (cell death) module is transformed into a model of \emph{reasonable} size, namely 180 Boolean variables (model \verb|F5_RA_apoptosis_executable_module.sbml| of supplementary material S3, and model ``RA-apoptosis'' of Section~\ref{sec:eval}).
The study of such model, though, is a big hurdle.
Indeed, as stated in the article about another model of the same size:
\emph{``The size of the CaSQ-inferred MAPK model (181 nodes) made the calculation of stable states a non-realistic endeavour.''}

@@

However, using a state-of-the-art tool like PyBoolNet~\cite{klarner2015computing} on that model actually fails at the phase of prime-implicant generation.
And hence, it is not possible to extract any (complex) attractor at all.
This is also true for the Alzheimer model also mentioned in that same article and originally from~\cite{ogishima2016alzpathway} (\verb|F4| file in the original supplementary material, and ``Alzheimer'' in Table~\ref{tab:result_real}), but actually not for the MAPK model for which the first trap spaces can be obtained in reasonable time.
The current practice usually revolves then around fixing some inputs to plausible values and reducing the model accordingly.
While this approach makes sense, it relies on potentially arbitrary decisions, and \emph{hides away} critical modelling choices that were actually not part of the original Boolean model or even of the starting map.

@@

To assess the efficiency of the proposed method, we compare it with the state-of-the-art method implemented in the tool PyBoolNet~\cite{klarner2015computing,klarner2017pyboolnet} on both its own repository of models and large models from the literature.

@@

All the models and a CoLoMoTo notebook realizing the benchmarks can be found at \url{https://github.com/soli/trap-spaces-as-siphons}. These can be run on a Docker image in the cloud by clicking the ``Binder'' button.
